---
title: "IntroductionToR"
author: "Maja Kuzman"
date: "7/1/2020"
output: html_document
---

# R, Rstudio, Rmd

---
### How can we use R and RStudio?  

- console   
- script   
- notebook:   
     new chunk: ctrl + ALT + i 

--
### What can we do with R? 

-  We can use R as a calculator!  

### Try it:  
     
Add a new chunk (ctrl + ALT + i).   
Calculate the result of 13%/%3 and 13%%3. 
( 13/3 = 4*3+1 )


*We can even do better than that...* 


# Variables and data structures

### Variables 


```{r}
myFirstNumber <- 0.1
myFirstVector <- c(2, 3, 7, 8)

```

-> call the variable by name  
-> print() function  
-> one line, part of code shaded and CTRL +ENTER

```{r}
print(myFirstNumber)
myFirstNumber

```


### Data structures in R  

vectors  
matrix  
data.frame  
list
factors  


# Vectors

## The basics

### What we can do with vectors:  

#### Make a vector c()  

You can make a vector by using the function c() (concatenate). Here is an example of vectors myFirstvector, and myFirstSequence:

```{r}
myFirstVector <- c("some words","p","word", "last one")
myFirstSequence <- 1:4
```

---

## Subsetting [ ]

Print the whole vector

```{r}
myFirstVector
myFirstSequence
```

Print third element in a vector
```{r}
myFirstVector[3]
myFirstSequence[3]
```

---

### Access multiple elements:  

1) BY POSITION  

Provide A VECTOR of positions to look at:  
```{r}
myFirstVector
```

THIS WON'T WORK:
```{r, eval=FALSE}
myFirstVector[1,3]
```

```{r}
myFirstVector[c(1,3)]
```

```{r}
somePositions <- c(1,3)
somePositions
```

```{r}
myFirstVector[somePositions]
```

---

### Access multiple elements:  

2) BY INCLUSION  

Provide a LOGICAL vector :  
```{r}
myFirstVector
myFirstVector[c(TRUE, FALSE, TRUE,FALSE)]
```

```{r}
someLogicalVector <- c(TRUE,TRUE, FALSE, FALSE)
someLogicalVector
```

```{r}
myFirstVector[someLogicalVector]
```


---

### Exercise! 

1. Create a vector named myvector that contains numbers 15,16,17,18 and 20.  
2. Get 2nd and 4th number in the vector by subsetting.  
3. Get 2nd and 4th number in the vector by providing a logical vector.  


Solution: 
CTRL+Alt+i will insert a new chunk  


---

### Think about it!

#### What happened here??:  

```{r}
someothervector <- c(1,0,1,0,1)
myFirstVector
myFirstVector[someothervector]
myFirstVector[as.logical(someothervector)]
```

--

### Basic operation on vectors  

Same as on numbers:  
    +  
    -  
    /  
    *  

Example:   

#### Multiplication  by constant  

```{r}
someothervector
someothervector * 0.5
```

---

#### Multiplication by other vector:

I) SAME size  

```{r}
someothervector
1:5
someothervector * 1:5
```

II) DIFFERENT size : recycling *because why not.*

```{r}
someothervector
c(0.3,0.1)
someothervector*c(0.3,0.1)
# c(1,0,1,0,1)* c(0.3, 0.1, 0.3, 0.1,0.3, 0.1)
c(1,2,3,4,5,6) * 1:2 # doesnt produce a warning c(1,2,3,4,5,6) * c(1,2, 1,2, 1,2)
```

---

### Basic comparisons  

The following will return a logical vector for every compared position:

```{r}
someothervector
someothervector == 1
```

```{r}
someothervector == c(1,0,1,0,1)
```

```{r}
someothervector
someothervector>0
```

--

#### what happened here?  

```{r}
someothervector[someothervector>0]

```
*remember: subsetting by logical indices returns ELEMENTS on POSITIONS WHICH ARE TRUE*

---
### Exercise: Multiplication, recycling and comparison.   
1. Multiply your myvector by c(0.1, 0.2)  
     -> what do you expect to get??
2. Check if you get what you expected by comparing it to vector you expect to get :)   


Solution:  
(get new chunk by ctrl+alt+i)


---

### Exercise: Multiplication, recycling and comparison.   
1. Return only ELEMENTS in myvector that are smaller then 17.
     -> hint - use subsetting and comparison. 


```{r, eval=FALSE}
myvector[ ??? ]

```
### Logical operators

AND: & 
OR: |  
NOT: ! 

They are used in a following way: 

```{r, eval=FALSE}
firstLogical <- c(TRUE, TRUE, FALSE, FALSE) 
secondLogical <- c(TRUE, FALSE, TRUE, FALSE)
firstLogical & secondLogical
firstLogical | secondLogical
! firstLogical
```


### Exercise: Subset by logical indexes

remember:  
     x == 0 - where is x equal to 0  
     x%%5 -> gives you the modulo while dividing x by 5  
     x != 0 - not equal to   

1. Return all the elements from your myvector that are divisable by 3.  


extra:
2. Return all the elements from your myvector which are on EVEN positions (2,4,6,...)  


## Functions 
#### Some useful functions  

---

### length - Gives you the length of the vector:    

```{r}
length(someothervector)
```

--

### unique -Gives you all unique elements in your vector:   
```{r}
unique(someothervector)
```

### table : gives you list of all elements and counts them  

```{r}
table(someothervector)
```

---
## Math:  


```{r}
sum(someothervector)
mean(someothervector)
sd(someothervector)
summary(someothervector)
```


### HELP!!??!!

```{r, eval=FALSE}
?runif
example(runif)
```

